---
title: 'Final Project Heart Attack Predictions'
author: "Darcey Ferguson"
date: "10/28/2025"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
heart_data <- read.csv(file = "heart_attack_dataset_age.csv", header = TRUE)
```

# **ABSTRACT**

Coming soon to DVD

# **BACKGROUND**

Every year about 805,000 Americans have a heart attack. (Tsao et al., 2023) A heart attack can pose pain from a mild discomfort to severe crushing pain. (Kulkarni, 2025) This health risk has become like a ticking bomb being able to strike anyone. With this disease that has the ability to strike a variety of different individuals, research has begun to try and predict risk factors or conditions that can cause an individual to be more susceptible to having a heart attack. Long-term or chronic stress can lead to higher levels of inflammation in the body which leads to more plague buildup in the arteries. (Katella, 2024) Additionally, stress also causes hormones such as adrenaline. (Katella, 2024) Adrenaline increases mental alertness and the heart beats faster and raises blood pressure. (Katella, 2024) Prolonged stress leads to heart damage. 

With this in mind we turn to what might cause an individual to have prolonged stress. Specifically, undergraduate students often experience stress related to adapting to higher education and managing coursework. (Pérez-Jorge, 2025) Postgraduate students also exhibit stress about advanced research, thesis completion, and balancing academic work and professional responsibilities. (Pérez-Jorge, 2025) When it comes to pursuing higher education there are a lot of stressful responsibilities for individuals. This could be a cause of having a risk of heart attacks because of the impact of the stress on the body. 

Another factor that might contribute to a higher risk of heart attacks is gender. Researchers have found that men are twice as likely to have a heart attack than women. (Harvard Health Publishing, 2016) There is also some suspicion that hormones in women before menopause cause more defense for women against heart attacks. (Harvard Health Publishing, 2016) 

Knowing this information brings a lot of question of how much factors like stress, gender, or education could lead to an individual being more likely to have a heart attack. 

# **STUDY QUESTIONS and HYPOTHESIS**

## **Questions**  

Does the gender, education level, and/or stress level an individual has affect the probability of them getting a heart attack?

## **Hypothesis**
  
If an individual is male, has a higher level of education, and a higher stress level then we will see statistical significance with correlation to having a heart attack because of how these factors affect heart conditions over time.
  
## **Prediction**

An individual is male, has a higher level of education, and a higher the stress level they will be more likely to have a heart attack at some point in their life

# **METHODS**

The database chosen for this study was the Kaggle database, and the heart attack prediction set was the one chosen for this study. It was chosen because it had enough entries for our purposes and many variables to test. Once the variables were chosen, an excel spreadsheet was created to hold all the data. Multiple linear regression models were made testing variables against the outcomes of heart attack. After all the variables were tested individually, they were then all tested at the same time vs heart attack outcome to test for a correlation between all of them.

## **Heart Attack vs Stress Level**

```{r Heart Attack vs StressLevel Violin Plot, eval=FALSE}
ggplot(heart_data, aes(x = Outcome, y = StressLevel, fill = Outcome)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", alpha = 0.5) +
  labs(
    title = "Stress Levels by Heart Attack Outcome",
    x = "Heart Attack Outcome",
    y = "Stress Level (1–9)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none"
  )

```

## **Heart Attack vs Stress Level**

```{r Stress vs Heart Attack Ridge Plot, eval=FALSE }
library(ggplot2)
library(ggridges)

ggplot(heart_data, aes(x = StressLevel, y = Outcome, fill = Outcome)) +
  geom_density_ridges(alpha = 0.7, color = "black", scale = 1) +
  scale_fill_manual(values = c("No Heart Attack" = "skyblue", "Heart Attack" = "tomato")) +
  labs(
    title = "Distribution of Stress Levels by Heart Attack Outcome",
    x = "Stress Level (1–9)",
    y = "Heart Attack Outcome",
    fill = "Outcome"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none",
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```
## **Interpretation of Stress Level vs Heart Attack Plot**

Coming soon to DVD

## **Gender vs Heart Attack Heat Plot**

```{r Gender vs Heart Attack Heat Plot, eval=FALSE}
library(dplyr)
library(ggplot2)
library(viridis)  

# Summarize counts
gender_counts <- heart_data %>%
  group_by(Gender, Outcome) %>%
  summarise(count = n(), .groups = "drop")

#Heat map
ggplot(gender_counts, aes(x = Gender, y = Outcome, fill = count)) +
  geom_tile(color = "white", size = 0.8) +  # white borders between tiles
  geom_text(aes(label = count), color = "white", size = 6, fontface = "bold") +
  scale_fill_viridis(option = "C", direction = -1) +  # smooth and modern gradient
  labs(
    title = "Heart Attack Outcomes by Gender",
    x = "Gender",
    y = "Heart Attack Outcome",
    fill = "Count"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 20),
    axis.text = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 16),
    legend.title = element_text(face = "bold", size = 14),
    legend.text = element_text(size = 12)
  )
```
## **Interpretation of Gender vs Heart Attack Plot**

Coming soon to DVD

## **Education Level vs Heart Attack**

```{r Heart Attack Vs Education Level lollipop Plot, eval=FALSE}
library(dplyr)
library(ggplot2)

# Summarize counts
edu_counts <- heart_data %>%
  group_by(EducationLevel, Outcome) %>%
  summarise(count = n(), .groups = "drop")

# Fancy lollipop chart
ggplot(edu_counts, aes(x = EducationLevel, y = count, color = Outcome)) +
  geom_segment(aes(x = EducationLevel, xend = EducationLevel, y = 0, yend = count), linewidth = 1.2) +
  geom_point(size = 6) +
  scale_color_manual(values = c("No Heart Attack" = "skyblue", "Heart Attack" = "tomato")) +
  labs(
    title = "Heart Attack Outcomes by Education Level",
    x = "Education Level",
    y = "Count of Individuals",
    color = "Heart Attack Outcome"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```
## **Interpretation of Education Level vs Heart Attack**

Coming soon to DVD

## **All Variable Combined Plot**

```{r, Stress Levels, Education Level, Gender, Heart Attack, eval=FALSE}
library(ggplot2)

ggplot(heart_data, aes(
  x = EducationLevel,
  y = StressLevel,
  color = Outcome,       
  alpha = Outcome,  
  shape = Gender
)) +
  geom_jitter(width = 0.3, size = 4) +
  scale_alpha_manual(values = c("No Heart Attack" = .9, "Heart Attack" = 1)) +
  scale_shape_manual(values = c("Female" = 1, "Male" = 16)) +  # hollow vs filled
  scale_color_manual(values = c("No Heart Attack" = "lightblue3", "Heart Attack" = "indianred2")) + 
  labs(
    title = "Stress Levels by Education, Gender, and Heart Attack Outcome",
    x = "Education Level",
    y = "Stress Level (1–9)",
    color = "Heart Attack Outcome",
    alpha = "Heart Attack Outcome",
    shape = "Gender"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

```

## **Interpretation of All Variables Combined Plot**

Coming soon to DVD

# **DISCUSSION**

Coming soon to DVD

# **CONCLUSION**

Coming soon to DVD

# **REFERENCES**   

Coming soon to DVD